Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2014-07-28 concepts.tex
AA: prima parte finalizzata, manca solo di una lettura complessiva
Some final touches, added more results comment

\begin{figure}[t]
\label{fig:psos}
\includegraphics[width=0.5\columnwidth]{img/Pareto_set_and_front}
\caption{Relation between Pareto-set and Pareto front.}
\end{figure}

@@

In the design process, there are often a lot of ``local decisions''
in which the designer can take advantage of well-known results or
heuristics. For example, if we consider a VLIW architecture in a configuration with very few registers, it is well-known that increasing the number of ALUs does not bring tangible better performances
but only cause an increase in area occupation. An intelligent exploration
algorithm would not waste time evaluating configurations with many
ALUs and few registers. If $p_{1}$ is the parameter ``number of
registers'' and $p_{2}$ is the parameter ``number of ALUs'', we
can say that the region:
\[
R=\left\{ \left.\left(v_{1},v_{2}\right)\in P_{1}\times P_{2}\right|v_{1}<s_{1},v_{2}>s_{2}\right\} 
\]
 (where $s_{1}$ and $s_{2}$ are some threshold values) is \emph{uninteresting}. This clearly shows that there are cases when, if a parameter lies in
a certain range, there are ranges of the other one that are not worth being explored. Intuitively, the cartesian product of these ranges is an \emph{uninteresting region}.

@@

	\begin{figure}[t]
	\includegraphics[width=0.5\columnwidth]{img/small_and_big}
	\figLC{small_and_big}{In this example 5 configurations
	are evaluated in a small region and in a bigger region. In can be
	seen that smaller region is more crowded.}
	\end{figure}

@@

	\begin{figure}[h]
		\includegraphics[width=0.5\columnwidth]{img/contiguous_regions}
		\figLC{contiguous_regions}{Contiguous and non contiguous regions. \comment{Inserire a dx una figura con le contiguous regions mergiate.}	}
	\end{figure}

>>>>

\begin{figure}[t]
\label{fig:psos}
\center
\includegraphics[width=0.5\columnwidth]{img/Pareto_set_and_front}
\caption{Relation between Pareto-set and Pareto front.}
\end{figure}

@@

In the design process, there are often a lot of ``local decisions'' in
which the designer can take advantage of well-known results or
heuristics. For example, if we consider a highly parallel
computer architecture in a configuration with very few registers, it is
well-known that increasing the number of ALUs does not bring tangible
better performances but only cause an increase in area occupation. An
intelligent exploration algorithm would not waste time evaluating
configurations with many ALUs and few registers. If $p_{1}$ is the
parameter ``number of registers'' and $p_{2}$ is the parameter
``number of ALUs'', we can say that the region:
\[
R=\left\{ \left.\left(v_{1},v_{2}\right)\in P_{1}\times P_{2}\right|v_{1}<s_{1},v_{2}>s_{2}\right\} 
\]
 (where $s_{1}$ and $s_{2}$ are some threshold values) is \emph{uninteresting}. This clearly shows that there are cases when, if a parameter lies in
a certain range, there are ranges of the other one that are not worth being explored. Intuitively, the cartesian product of these ranges is an \emph{uninteresting region}.

@@

	\begin{figure}[t]
	\center
	\includegraphics[width=0.5\columnwidth]{img/small_and_big}
	\figLC{small_and_big}{In this example 5 configurations
	are evaluated in a small region and in a bigger region. In can be
	seen that smaller region is more crowded.}
	\end{figure}

@@

	\begin{figure}[t]
		\center
		\includegraphics[width=0.8\columnwidth]{img/contiguous_regions}
		\figLC{contiguous_regions}{Contiguous and non contiguous regions.}
	\end{figure}

### 2014-07-28 experiments.tex
Some final touches, added more results comment

\section{A Case Study: VLIW Hardware/Software Design Space}
\label{sec:ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we will be using a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The same philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item{Multi-Objective trade-offs}: VLIW architectures let designers
easily trade between power, energy and performances, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space space representation
\item{Compiler awareness}: Instruction Level Parallelism (ILP)
is statically obtained by the compiler which
performs code transformations being aware of the underlying hardware
configuration.  This tight hardware/software dependence, makes the
VLIW scenario perfectly suitable for investigating hardly predictable
hardware/software interactions.
\end{itemize}

@@

Next, a set of the most impacting code compilation parameters has been
selected and included in the configuration space:
\begin{itemize}
\item{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item {max\_unroll\_allowed}: The number of unroll iterations allowed
\item{regroup\_only}: Avoids inlining 
\item{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal 
\item{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
do\_modulo\_scheduling  Modular scheduling 
\item{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

@@

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 500
  configurations.\label{fig:pareto_fronts_500}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{tabular}}
\end{table}
Low budget Pareto sets are shown in
Figure~\ref{fig:pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly depended upon the application being considered.
These results are obtained in about 5 eras, and the main aim was
demonstrating that with a so limited budget PS still can compete with
genetic approach while even discovering very different Pareto fronts.
Continuing with a more extended simulation budget as depicted in
Figure~\ref{fig:pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc~\ldots.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective. A
comparison between the variation range for different benchmarks
between a PS and a GA exploration for both power dissipation and
execution time is shown in Fig.~\ref{fig:dispersion}(a) and (b)
respectively. As it can be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than that provided by a
MOGA exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range for different benchmarks between PS and GA
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

@@

Fig.~\ref{fig:dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA still shows a tendency in
discovering more uniformly dispersed points (i.e. higher values). These results,
together with the variation ranges of Fig.~\ref{fig:dispersion}(a) and
(b) confirm the the availability of density vs extension trade-offs
when comparing PS and multi-objective genetic approaches.

>>>>

\section{A Case Study: VLIW Hardware/Software Design Space}
\label{sec:ee}
\subsection{Motivation}
In order to evaluate the behavior of the proposed approach
we will be using a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The same philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures let designers
easily trade between power, energy and performances, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

@@

Next, a set of the most impacting code compilation parameters has been
selected and included in the configuration space:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal 
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
do\_modulo\_scheduling  Modular scheduling 
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

@@

\begin{table}
  \tbl{Pareto fronts found by PS and GA for a fixed budget of 500
  configurations.\label{fig:pareto_fronts_500}}{
  \begin{tabular}{ccc}
    \includegraphics[width=0.30\textwidth]{pictures/alloca_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/bmm_500.eps} & 
    \includegraphics[width=0.30\textwidth]{pictures/fir_int500.eps} \\
    \includegraphics[width=0.30\textwidth]{pictures/mm_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/sha_500.eps} &
    \includegraphics[width=0.30\textwidth]{pictures/wave_500.eps} 
  \end{tabular}}
\end{table}
Low budget Pareto sets are shown in
Figure~\ref{fig:pareto_fronts_100}. With a so small number of
configuration being investigated the two sets are often overlapping,
being also strongly depended upon the application being considered.
These results are obtained in about 5 eras, and the main aim was
demonstrating that with a so limited budget PS still can compete with
genetic approach while even discovering very different Pareto fronts.
Considering the extended simulation budget as depicted in
Figure~\ref{fig:pareto_fronts_500}, we can qualitatively observe that
even if PS does not strictly outperforms genetic in terms of dominance, it shows in
some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behaviour as
``better'' or ``worse'' as compared to the genetic one, it depends on
various design factors such as desired results granularity, prefixed objective
range constraints, estimated error of the models used etc~\ldots.

In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. For a given
objective, the variation range represents the ratio between the
maximum and the minimum value observed for that objective. A
comparison between the variation range for different benchmarks
between a PS and a GA exploration for both power dissipation and
execution time is shown in Fig.~\ref{fig:dispersion}(a) and (b)
respectively. As it can be observed, the PS exploration provides solutions which fall on a
range that is, on average, wider than that provided by a
GA exploration for power dissipation and execution time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range for different benchmarks between PS and GA
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

@@

Fig.~\ref{fig:dispersion}(c) shows the average normalised absolute
dispersion errors for different benchmarks for PS and GA
exploration. As it can be observed, GA still shows a tendency in
discovering more uniformly dispersed points (i.e. higher values in the
chart). These results,
together with the variation ranges of Fig.~\ref{fig:dispersion}(a) and
(b) confirm the the availability of density vs extension trade-offs
when comparing PS and GA approaches. Again, it
should be pointed out how a definitive claim about the superiority of
the PS behaviour is stricly dependent on the design 
requirements and constraints, e.g. whether we need wide or dense
results. Nevetheless, the proposed algorithm introduces a new
perspective on thinking the allocation of a simulations budget, paving the way for
the development of further strategies for addressing the problem of
multi-objective hardware/software desing space exploration.

### 2014-07-28 introduction.tex
Some final touches, added more results comment

In these cases a Design Space
Exploration (DSE) strategy is required, i.e. an algorithm that selects
only a tractable subset of all possible configurations.
The final result of a design space exploration is a subset of the
configurations called \emph{Pareto set} (\cite{pareto}, see also
\defR{pareto-set}). The corresponding values of the objectives for
each configuration of the Pareto Set is called \emph{Pareto front}:
it represents the trade-offs between objectives, i.e.  for each point
of the Pareto front there is no other configuration performing better
for all the objectives considered. Of course this is only a first
high-level system simulation, thus these few promising
configurations will need to be further evaluated in a next step of the design
flow using a more accurate low level simulation (e.g. register
transfer level).

@@

Some exploration strategies assume some kind of knowledge about the
system parameters, their meaning and their impact on design
objectives.  The \emph{Dependency analysis}, proposed in
\cite{givargis_tvlsi02}, is meant to take advantage of the parameter
dependencies. Starting from them, he can construct a ``dependency graph''
and recognize clusters, i.e. subsets of strongly dependency-connected
parameters. Each cluster is exhaustively evaluated and its ``local''
Pareto set is found. Then, all local Pareto sets are merged. In this
way, a series of local exhaustive evaluations are performed instead of
an exhaustive evaluation of all the possible configurations. Some
problems arise: i) In real scenarios, it may be difficult to recognize
really independent clusters of parameters, because there may be
interdependencies among a large number of parameters. This may lead to
the need of an evaluation of almost all the possible configurations.
ii) A designer may not have a precise and complete a-priori picture of
the dependencies among parameters; for this reason the need of
``automated approaches for computing interdependencies'' is declared
in the same paper.  These drawbacks are not present in our algorithm,
since, although it leverages dependencies as in
\cite{givargis_tvlsi02}, it is able to detect them automatically, with
no a-priori knowledge required. Moreover, our algorithm is able to
capture also ``local dependencies", i.e. dependencies that emerge only
among certain ranges of parameters and may not hold when considering
the entires ranges. This cannot be modelled in Dependency Analysis.

>>>>

In these cases a Design Space
Exploration (DSE) strategy is required, i.e. an algorithm that selects
only a tractable subset of all possible configurations.
The final result of a design space exploration is a subset of the
configurations called \emph{Pareto set} (\cite{pareto}, see also
\defR{pareto-set}). The corresponding values of the objectives for
each configuration of the Pareto Set is called \emph{Pareto front}:
it represents the trade-offs between objectives, i.e.  for each point
of the Pareto front there is no other configuration performing better
for all the objectives considered (see Figure~\ref{fig:psos}). Of course this is only a first
high-level system simulation, thus these few promising
configurations will need to be further evaluated in a next step of the design
flow using a more accurate low level simulation (e.g. register
transfer level).

@@

Some exploration strategies assume some kind of knowledge about the
system parameters, their meaning and their impact on design
objectives.  The \emph{Dependency analysis}, proposed in
\cite{givargis_tvlsi02}, is meant to take advantage of the parameter
dependencies. Starting from them, he can construct a ``dependency graph''
and recognize clusters, i.e. subsets of strongly dependency-connected
parameters. Each cluster is exhaustively evaluated and its ``local''
Pareto set is found. Then, all local Pareto sets are merged. In this
way, a series of local exhaustive evaluations are performed instead of
an exhaustive evaluation of all the possible configurations. Some
problems arise: i) In real scenarios, it may be difficult to recognize
really independent clusters of parameters, because there may be
interdependencies among a large number of parameters. This may lead to
the need of an evaluation of almost all the possible configurations.
ii) A designer may not have a precise and complete a-priori picture of
the dependencies among parameters; for this reason the need of
``automated approaches for computing interdependencies'' is declared
in the same paper.  These drawbacks are not present in PS,
since, although it leverages dependencies as in
\cite{givargis_tvlsi02}, it is able to detect them automatically, with
no a-priori knowledge required. Moreover, our algorithm is able to
capture also ``local dependencies", i.e. dependencies that emerge only
among certain ranges of parameters and may not hold when considering
the entires ranges. This cannot be modelled in Dependency Analysis.
